Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-07-23 content.tex
Reviewed

Here we introduce a method to quantify the importance of a novel tree prior.
The method starts with a phylogeny generated by the new model for the tree prior. Next, nucleotide sequences are simulated on this phylogeny. Then, using the tree priors built-in into BEAST2, a Bayesian posterior distribution of phylogenies is inferred. We can then compare the inferred and simulated phylogenies. How to properly perform this comparison forms the heart of our method.
Only new diversification models that result in a large discrepancy between inferred and simulated phylogenies will be worth the effort and computational burden 
to implement a species tree prior for in a Bayesian framework.

@@

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of a (DNA) root sequence, a (DNA) mutation rate, a clock model and a nucleotide substitution
model, which will be referred to as site model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported site models are JC, HKY, TN and GTR. Only the strict
clock model is currently supported in this step.

@@

We show the usage of \verb;pirouette; by gradually introducing
its features.
First, to get an idea of the baseline error, 
we measure the Bayesian inference error when we start from a phylogeny 
generated under a known and standard tree prior.
Second, to establish that the true tree prior is indeed the best, 
we also measure the inference error made adopting a
best candidate inference model.
Lastly, we quantify the impact of the tree prior in the Bayesian inference. 
We do so by separating the baseline error from the complete error 
running the pipeline starting from a tree generated by an unknown 
or non-standard tree prior.

@@

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's 
inference by measuring the discrepancy between a given tree and a posterior 
distribution of phylogenies obtained as result of the inference process. 
Due to stochasticity, posterior trees will generally differ from the given 
phylogeny $\tau_{\mathit{G}}$, even when the generative tree 
prior $\mathit{p_{G}}$ and alignment model $\mathit{A}$ are the same as those 
in the inference model $\mathit{I}$.
Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' 
all the tree priors that are available to be used within an inference model (see Table~\ref{tab:options}).

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will demonstrate it in section
 \ref{Comparing to background noise}.

@@

We start from the tree generated by a non-standard (and unknown) tree prior, having six taxa and a crown age of ten:
\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/true_tree.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
  }
\end{figure}

Most of the other settings are the same as before: we re-use the alignment parameters (Listing~\ref{lst:create_alignment_params}), as well as the experiments (Listing~\ref{lst:create_all_experiments}).
This time, however, we enable twinning (see Subsection~\ref{subsec:twinning}),
by creating twinning parameters.
Creating these parameters is trivial with the default settings.
For clarity, however, we explicitly show the most important arguments:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "birth_death",
  method = "random_tree"
)
\end{lstlisting}

@@

Figure~\ref{fig:example_3} illustrates the primary result of our pipeline: it shows the error distributions for the true tree and the twin tree when either the generating model or the best candidate model is used. For this specific case it does not seem to matter that the site and clock models used in inference are different from those used in generating the data, but the difference between the error distributions for the true tree and the twin tree suggest that the choice of tree prior does matter.

We note, however, that all examples used only one original tree,
where any speciation process produces a whole range of trees.
One tree is not enough to determine the impact of a tree prior on Bayesian inference.
However, if the same procedure were repeated and performed on a distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the 
quality of the inference.
Also a twin tree does not always result in a lower error distribution,
as the stochasticity in generating a twin tree will occasionally
yield a tree of very low probability.

In conclusion, \verb;pirouette; can show the errors
when adopting a new generative prior
in comparison to the errors made by a standard prior.
The user can then judge whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this is indeed the case, one should implement 
the novel tree prior as an addition to 
the favorite Bayesian inference tool.

@@

\verb;pirouette;'s development takes place on GitHub,
\url{https://github.com/richelbilderbeek/pirouette},
which allows submitting bug reports, requesting features, 
or adding code. To ensure a high quality, \verb;pirouette; 
uses a continuous integration service, has a code coverage of above 95\%
and enforces the most commonly used R style guide (\cite{style_guide}).

>>>>

Here we introduce a method to quantify the importance of a novel tree prior.
The method starts with a phylogeny generated 
by the new model. 
Next, nucleotide sequences are simulated on this phylogeny. 
Then, using the tree priors built-in into BEAST2, 
a Bayesian posterior distribution of phylogenies is inferred. 
We then compare the inferred and simulated phylogenies. 
How to properly perform this comparison forms the heart of our method.
Only new diversification models that result 
in a large discrepancy between inferred and simulated phylogenies 
will be worth the effort and computational burden 
to implement a species tree prior for in a Bayesian framework.

@@

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of 
a (DNA) root sequence, a (DNA) mutation rate, a clock model 
and a nucleotide substitution model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported nucleotide substitution model, which
we will refer to as site models, 
are JC, HKY, TN and GTR. Only the strict
clock model is currently supported in this step.

@@

We show the usage of \verb;pirouette; by gradually introducing
its features.
First, to get an idea of the baseline error, 
we measure the Bayesian inference error when we start from a phylogeny 
generated under a known and standard tree prior.
Second, to establish that the true tree prior is indeed the best, 
we also measure the inference error made by a
best candidate inference model.
Lastly, we quantify the impact of the tree prior in the Bayesian inference. 
We do so by separating the baseline error from the complete error 
running the pipeline starting from a tree generated by an unknown 
or non-standard tree prior.

All the figures shown in this section are shown as-is, without any
aesthetical modifications..

@@

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's 
inference by measuring the discrepancy between a given/true tree and a posterior 
distribution of phylogenies obtained as result of the inference process. 
Due to stochasticity, posterior trees will generally differ from the given 
phylogeny $\tau_{\mathit{G}}$, even when the generative tree 
prior $\mathit{p_{G}}$ and alignment model $\mathit{A}$ are the same as those 
in the inference model $\mathit{I}$.
Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' 
all the tree priors that are available 
to be used within an inference model (see Table~\ref{tab:options}).

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will demonstrate it in section
\ref{Comparing to background noise} below.

@@

We start from the tree generated by a non-standard (and unknown) tree prior, 
having six taxa and a crown age of ten:
\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/true_tree.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
    The scale overlaps with the lower branch, which is the default 
    behavior for the plotting function used.
  }
\end{figure}

Most of the other settings are the same as before: 
we reuse the alignment parameters (Listing~\ref{lst:create_alignment_params}), 
as well as the experiments (Listing~\ref{lst:create_all_experiments}).
This time, however, we enable twinning (see Subsection~\ref{subsec:twinning}),
by creating twinning parameters.
Creating these parameters is trivial with the default settings.
For clarity, however, we explicitly show the most important arguments:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "yule",
  method = "random_tree"
)
\end{lstlisting}

@@

Figure~\ref{fig:example_3} illustrates the primary result of our pipeline: 
it shows the error distributions for the true tree and the twin tree 
when either the generating model or the best candidate model is used. 
The clear difference between the error distributions 
for the true tree and the twin tree suggest 
that the choice of tree prior does matter.

We note, however, that all examples used only one original tree,
where any speciation process produces a whole range of trees.
One tree is not enough to determine the impact 
of a tree prior on Bayesian inference.
However, if the same procedure were repeated and 
performed on a distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the 
quality of the inference.
Also a twin tree does not always result in a lower error distribution,
as the stochasticity in generating a twin tree will - with
very low probability - yield a tree of that same very low probability.

In conclusion, \verb;pirouette; can show the errors
when adopting a new generative prior
in comparison to the errors made by a standard prior.
The user can then judge whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this is indeed the case, one is warranted to implement 
the novel tree prior as an addition to 
his/her favorite Bayesian inference tool.

@@

\verb;pirouette;'s development takes place on GitHub,
\url{https://github.com/richelbilderbeek/pirouette},
which allows submitting bug reports, requesting features, 
and adding code. To ensure a high quality, \verb;pirouette; 
uses a continuous integration service, has a code coverage of above 95\%
and enforces the most commonly used R style guide (\cite{style_guide}).

### 2019-07-23 supplement.tex
Refresh data
Reviewed

@@

\input{example_1/esses.latex}

>>>>

These supplementary materials show the full pipeline and
added diagnostics for the examples in the main article.

As in the main article, all the figures shown in this section are shown as-is, 
without any aesthetical modifications, with the exception that the arrangement
of the sub-figures (for example, aligning true and twin tree
horiztonally) is done manually.

@@

\input{example_1/esses_gen.latex}
